Automated Formal Synthesis of Digital Controllers for State-Space Physical Plants

### 2017-01-24 cav2017/cav2017.tex
replaced "ise" by "ize"

We describe and evaluate two approaches for synthesizing digital controllers 
for state-space physical plants. Both approaches consider a counterexample 
guided inductive synthesis (CEGIS) loop. In the first approach, 
we devise a digital controller that stabilises the system; then, we verify its 
potential safety by unfolding the dynamics of the system and checking for 
a completeness threshold. In the second approach, we employ abstract acceleration
to accelerate the dynamics forward instead of symbolically unwinding the 
transition relation~\cite{cattaruzza2015unbounded}. We provide experimental results 
showing that both approaches are able to efficiently synthesize stable and safe controllers 
for a set of intricate physical plant models taken from the digital 
control literature: the median runtime for our benchmark
set is \textcolor{red}{$XXX$}s, i.e., most controllers can be synthesized in less
than \textcolor{red}{$XXX$} seconds.

@@

Program synthesis is the problem of computing correct-by-design programs from high-level specifications; 
algorithms for this problem have made substantial progress in recent years.  
One such approach~\cite{itzhaky2010simple} inductively synthesises invariants to
generate the desired programs.

Program synthesizers are an ideal fit for synthesis of %parametric
digital controllers, since the semantics of programs capture effects such as FWL
precisely.  In~\cite{DBLP:conf/cdc/RavanbakhshS15}, the authors use CEGIS
for the synthesis of switching controllers for stabilizing continuous-time
plants with polynomial dynamics.  The work extends to affine systems, 
finding a major challenge in the hardness of solving linear arithmetic with the state-of-the-art SMT solvers.  
Since this approach uses switching models instead of linear dynamics for the digital controller, 
it entirely circumvents the FWL problem, but potentially incurs state-space explosion issues.  
This approach is not suitable for the kind of control we seek to synthesise.

@@

The work in \cite{DBLP:journals/corr/AbateBCCDKK16} investigates the synthesis of 
stabilizing controllers for continuous plants described as transfer functions, 
by exploiting bit-accurate verification of software-implemented 
digital controllers~\cite{Bessa16}.  While this work also make use of a CEGIS-based technique, 
the approach is restricted to synthesizing digital controllers for stable 
closed-loop systems described via transfer function models.  
In contrast, in the current paper we consider a state-space representation of the physical system, 
which is known to be more general \cite{Astrom08} and allows to synthesise controllers that
make the closed-loop system \emph{safe}.  Moreover, 
this work  integrates an abstraction/refinement step inside the main CEGIS loop.

The tool Pessoa~\cite{mazo2010pessoa} is a MatLab toolbox that synthesises correct-by-design
embedded control software. It is based on the description of a physical system
as an equivalent finite-state machine and on the computation of reachability properties over the finite-state 
machine. 
Based on this safety specification over state values, 
Pessoa can synthesise embedded controller software for a range of properties.  
%based on a safety specification over state values, as well as specification on characteristic functions of the controller. 
However, Pessoa does not have native inclusion of the ADC/DAC quantisation error needed for
synthesizing fixed-point digital controllers.

@@

The classical approach to control synthesis has often disregarded quantization, 
thus resulting in general to unsound digital controllers. 
Modern techniques focus on different elements of discretization effects, including delayed
response~\cite{Duggirala2015}, and Finite Word Length (FWL) semantics, 
with the goal either to verify (e.g.~\cite{daes20161}) or to optimise
(e.g.~\cite{oudjida2014design}) given implementations.

@@

Models \eqref{eq:ode} and \eqref{eq:plant} depend on external non-determinism in the form of input signals $u (t), u_k$, respectively. 
Feedback architectures can be employed to affect the properties and behaviours of the continuous process (the plant) of interest.   
%reduce the effect of parametric uncertainties and attenuate the effects of noise and disturbance. 
We are interested in the synthesis of digital feedback, 
as implemented on modern FPGAs or ICs. 
The most basic feedback architecture is the state feedback, 
where the control action $u_k$ (notice we work with the discretized signal) is computed by: 
%
\begin{equation}
\label{eq:controlaction}
u_k = r_{k} - K x_k. 
\end{equation}
%
Here $K \in \mathbb{R}^{m \times n}$ is a feedback gain matrix, 
and $r_{k}$ is a reference signal (again digital).   
%
The closed-loop model then takes the form 
\begin{align}
\label{eq:closedloopss}
x_{k+1} = ( A_d - B_d K ) x_k + B_d r_k.
\end{align}
The gain matrix $K$ can be set so that the closed-loop digitalised dynamics are shaped as desired, 
for instance according to a specific stability goal or around a specific dynamical behaviour \cite{astrom1997computer}. 
As argued later, in this work, we will target more complex objectives, such as quantitative safety requirements that are not typical in the digital control literature.

@@

We note that the effect on stability due to the fixed-word-length error is that the system may, instead of converging to the desired equilibrium point, converge to a point that is bounded away from the desired equilibrium point (see Appendix for details). 
In this paper, we are concerned with \emph{safe} controllers and therefore concerned only that the system converges to a point within the safety specification. Hence, it is sufficient to synthesise a stable controller based on infinite precision, and then verify safety.

@@

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Solution generalisation} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
Given the large solution space to be searched during the inductive
synthesis phase, we parametrise the target language, which induces a
lattice of progressively more expressive languages.  This enables us
to start by attempting to synthesise a program at the lowest point on
this lattice and increase the parameters of the language until we
reach a point at which the synthesis succeeds. Whenever we manage to
synthesise a candidate solution at some lower point on the lattice, we
must apply some form of solution generalisation to obtain a full
solution (i.e. a solution in the original search space).

As well as giving us an automatic search procedure, this
parametrisation greatly increases the efficiency of our system since
languages low down the lattice are very easy to decide safety for
(i.e. the validation oracle finds an answer fast). If a program can be
synthesised in a low-complexity language, the whole procedure finishes
much faster than if synthesis had been attempted in a high-complexity
language.

We investigate two different approaches of integrating solution generalisation 
with CEGIS:
\begin{itemize}
\item[1.] The first approach (described in Sec.~\ref{sec:CEGIS-precision-incrementation})
  uses a multi-staged verification phase and performs solution generalisation based on
the plant precision, the number of loop unrollings, 
and the sampling rate. Essentially, it finds a controller for an individual initial state
and input with a bounded time horizon and generalises to all states,
inputs, and time horizons. The safety check is performed in this approach 
via unfolding of the transition system corresponding to the synthesised controller.

%% (with a final verification loop that verifies
%% the continuous system behaviour against the previously verified
%% discrete system bi-simulation)
\item[2.] The second approach (described in Sec~\ref{sec:CEGIS-abstraction-refinement})
  finds a controller for a continuous initial set of states and input
  set, over an abstraction of the continuous dynamics that conforms to
  witness proofs at specific times; it further generalises the
  solution over all times by refining the number of witnesses.  
  In order to perform the safety check, this approach uses abstract acceleration
  \cite{cattaruzza2015unbounded} to compute the controller's set of states. 
  %\todo{clarify use of abstract acceleration. add references.}
\end{itemize}

@@

The controller synthesis is described in Fig.~\ref{fig:CEGIS-precision-increment}.
One important observation is that, for this approach, we verify and synthesise 
a controller over $k$ time steps. We then compute a completeness threshold, $\overline{k}$,
for this controller, and verify the controller correctness to $\overline{k}$ time steps.
Essentially, $\overline{k}$ represents the number of iterations 
required to sufficiently unwind the closed-loop state-space system 
such that the boundaries are not violated for any $k{>}\overline{k}$
(i.e. it is sufficient to unwind the closed-loop state-space system up to  
$\overline{k}$ in order to ensure that $\phi_{safety}$ holds).

@@

%\todo{please eventually itemize next}
\begin{enumerate}
\item The inductive synthesis phase (i.e. {\sc synthesise}) uses BMC to
compute a candidate solution $K$
that satisfies both the stability criteria detailed in 
Sec.~\ref{ssec:stability} and the safety specification in Sec.~\ref{ssec:safety}.
To synthesise a controller that satisfies the stability criteria, we require that
a computed characteristic polynomial satisifies Jury's criteria. The details of this
calculation can be found in the Appendix.

%% The stability criteria are satisfied if the characteristic polynomial of $A_d - B_d * K_d$ meets Jury's criteria. This is 
%% independent of the initial states. 
Regarding the latter, we synthesise a safe controller by unwinding the transition system
$k$ steps and picking a controller $K$ and an initial state such that the states at each step do
not violate the safety criteria. This is sound if the current $k$ is below the
completeness threshold, which we assume to be true in the {\sc synthesise} phase.
We also assume some
precision $\langle I_p,F_p\rangle$ for the plant and a sampling rate.  The
checks that all these assumptions are sound are performed by subsequent
{\sc verify} stages.

@@

\item The first {\sc verify} stage, {\sc safety}, checks that the candidate
solution $K$ is safe for all possible initial states, i.e., does not reach an unsafe
state within $k$ steps where again we assume $k$ to be under the completeness threshold. 
%The safety criteria here must be valid for all possible allowed initial states, and so we
After unwinding the transition system corresponding to the previously synthesised controller
$k$ steps, we check that the safety specification holds for any initial state.

@@

In the subsequent iteration, the synthesiser is unable to find further 
suitbale candidates and it returns UNSAT, meaning that the current precision is
insufficient.  Consequently, we increase the precision to $I_p=17$,
$F_p=7$.
%
Since the previous counterexamples were obtained at lower precision,
we remove them from the set of counterexamples.  Back in the {\sc
  synthesize} phase, we re-start the process with a candidate solution
with all coefficients $0$.  Next, the {\sc safety} verification stage
provides the first counterexample at higher precision, $x_0=[-0.5
  \ 0.5 \ 0.5]$ and {\sc synthesize} finds $K=[0 \ 0.01171875
  \ 0.015625]^T$ as a candidate that eliminates this counterexample.
However, this candidate triggers the counterexample
$x_0=[0.5\ -0.5\ -0.5]$ found again by the {\sc safety} verification
stage.  In the next iteration, we get the candidate $K=[0 \ 0
  \ -0.015625]$, followed by the counterexample $x_0 = [0.5 \ 0.5
  \ 0.5]$. Finally, {\sc synthesize} finds the candidate $K=[0.01171875
  \ -0.013671875 \ -0.013671875]$, which is validated as a final
solution by all verification stages.

@@

\item In this phase we synthesise a candidate controller

@@

%-------------------------------
\subsection{Threats to validity}
%-------------------------------
We report an assessment of both our approaches over a diverse set of real-world benchmarks. Nevertheless, this set of benchmarks is limited within the scope of this paper and the performance may not generalise to other benchmarks.

@@

The definition of asymptotic stability is that the system converges to a reference signal, in this case we use no
reference signal so an asymptotically stable system will converge to the origin. We know that the original system
with an infinite precision controller is stable, because we have synthesised it to meet Jury's criteria. Hence
$(A_d  - B_dK)^n*x_0$ must converge to zero.

>>>>

We describe and evaluate two approaches for synthesizing digital controllers 
for state-space physical plants. Both approaches consider a counterexample 
guided inductive synthesis (CEGIS) loop. In the first approach, 
we devise a digital controller that stabilizes the system; then, we verify its 
potential safety by unfolding the dynamics of the system and checking for 
a completeness threshold. In the second approach, we employ abstract acceleration
to accelerate the dynamics forward instead of symbolically unwinding the 
transition relation~\cite{cattaruzza2015unbounded}. We provide experimental results 
showing that both approaches are able to efficiently synthesize stable and safe controllers 
for a set of intricate physical plant models taken from the digital 
control literature: the median runtime for our benchmark
set is \textcolor{red}{$XXX$}s, i.e., most controllers can be synthesized in less
than \textcolor{red}{$XXX$} seconds.

@@

Program synthesis is the problem of computing correct-by-design programs from high-level specifications; 
algorithms for this problem have made substantial progress in recent years.  
One such approach~\cite{itzhaky2010simple} inductively synthesizes invariants to
generate the desired programs.

Program synthesizers are an ideal fit for synthesis of %parametric
digital controllers, since the semantics of programs capture effects such as FWL
precisely.  In~\cite{DBLP:conf/cdc/RavanbakhshS15}, the authors use CEGIS
for the synthesis of switching controllers for stabilizing continuous-time
plants with polynomial dynamics.  The work extends to affine systems, 
finding a major challenge in the hardness of solving linear arithmetic with the state-of-the-art SMT solvers.  
Since this approach uses switching models instead of linear dynamics for the digital controller, 
it entirely circumvents the FWL problem, but potentially incurs state-space explosion issues.  
This approach is not suitable for the kind of control we seek to synthesize.

@@

The work in \cite{DBLP:journals/corr/AbateBCCDKK16} investigates the synthesis of 
stabilizing controllers for continuous plants described as transfer functions, 
by exploiting bit-accurate verification of software-implemented 
digital controllers~\cite{Bessa16}.  While this work also make use of a CEGIS-based technique, 
the approach is restricted to synthesizing digital controllers for stable 
closed-loop systems described via transfer function models.  
In contrast, in the current paper we consider a state-space representation of the physical system, 
which is known to be more general \cite{Astrom08} and allows to synthesize controllers that
make the closed-loop system \emph{safe}.  Moreover, 
this work  integrates an abstraction/refinement step inside the main CEGIS loop.

The tool Pessoa~\cite{mazo2010pessoa} is a MatLab toolbox that synthesizes correct-by-design
embedded control software. It is based on the description of a physical system
as an equivalent finite-state machine and on the computation of reachability properties over the finite-state 
machine. 
Based on this safety specification over state values, 
Pessoa can synthesize embedded controller software for a range of properties.  
%based on a safety specification over state values, as well as specification on characteristic functions of the controller. 
However, Pessoa does not have native inclusion of the ADC/DAC quantisation error needed for
synthesizing fixed-point digital controllers.

@@

The classical approach to control synthesis has often disregarded quantization, 
thus resulting in general to unsound digital controllers. 
Modern techniques focus on different elements of discretization effects, including delayed
response~\cite{Duggirala2015}, and Finite Word Length (FWL) semantics, 
with the goal either to verify (e.g.~\cite{daes20161}) or to optimize
(e.g.~\cite{oudjida2014design}) given implementations.

@@

Models \eqref{eq:ode} and \eqref{eq:plant} depend on external non-determinism in the form of input signals $u (t), u_k$, respectively. 
Feedback architectures can be employed to affect the properties and behaviours of the continuous process (the plant) of interest.   
%reduce the effect of parametric uncertainties and attenuate the effects of noise and disturbance. 
We are interested in the synthesis of digital feedback, 
as implemented on modern FPGAs or ICs. 
The most basic feedback architecture is the state feedback, 
where the control action $u_k$ (notice we work with the discretized signal) is computed by: 
%
\begin{equation}
\label{eq:controlaction}
u_k = r_{k} - K x_k. 
\end{equation}
%
Here $K \in \mathbb{R}^{m \times n}$ is a feedback gain matrix, 
and $r_{k}$ is a reference signal (again digital).   
%
The closed-loop model then takes the form 
\begin{align}
\label{eq:closedloopss}
x_{k+1} = ( A_d - B_d K ) x_k + B_d r_k.
\end{align}
The gain matrix $K$ can be set so that the closed-loop digitalized dynamics are shaped as desired, 
for instance according to a specific stability goal or around a specific dynamical behaviour \cite{astrom1997computer}. 
As argued later, in this work, we will target more complex objectives, such as quantitative safety requirements that are not typical in the digital control literature.

@@

We note that the effect on stability due to the fixed-word-length error is that the system may, instead of converging to the desired equilibrium point, converge to a point that is bounded away from the desired equilibrium point (see Appendix for details). 
In this paper, we are concerned with \emph{safe} controllers and therefore concerned only that the system converges to a point within the safety specification. Hence, it is sufficient to synthesize a stable controller based on infinite precision, and then verify safety.

@@

%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
\subsection{Solution generalisation} 
%+++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++++
Given the large solution space to be searched during the inductive
synthesis phase, we parametrize the target language, which induces a
lattice of progressively more expressive languages.  This enables us
to start by attempting to synthesize a program at the lowest point on
this lattice and increase the parameters of the language until we
reach a point at which the synthesis succeeds. Whenever we manage to
synthesize a candidate solution at some lower point on the lattice, we
must apply some form of solution generalisation to obtain a full
solution (i.e. a solution in the original search space).

As well as giving us an automatic search procedure, this
parametrisation greatly increases the efficiency of our system since
languages low down the lattice are very easy to decide safety for
(i.e. the validation oracle finds an answer fast). If a program can be
synthesized in a low-complexity language, the whole procedure finishes
much faster than if synthesis had been attempted in a high-complexity
language.

We investigate two different approaches of integrating solution generalisation 
with CEGIS:
\begin{itemize}
\item[1.] The first approach (described in Sec.~\ref{sec:CEGIS-precision-incrementation})
  uses a multi-staged verification phase and performs solution generalisation based on
the plant precision, the number of loop unrollings, 
and the sampling rate. Essentially, it finds a controller for an individual initial state
and input with a bounded time horizon and generalizes to all states,
inputs, and time horizons. The safety check is performed in this approach 
via unfolding of the transition system corresponding to the synthesized controller.

%% (with a final verification loop that verifies
%% the continuous system behaviour against the previously verified
%% discrete system bi-simulation)
\item[2.] The second approach (described in Sec~\ref{sec:CEGIS-abstraction-refinement})
  finds a controller for a continuous initial set of states and input
  set, over an abstraction of the continuous dynamics that conforms to
  witness proofs at specific times; it further generalizes the
  solution over all times by refining the number of witnesses.  
  In order to perform the safety check, this approach uses abstract acceleration
  \cite{cattaruzza2015unbounded} to compute the controller's set of states. 
  %\todo{clarify use of abstract acceleration. add references.}
\end{itemize}

@@

The controller synthesis is described in Fig.~\ref{fig:CEGIS-precision-increment}.
One important observation is that, for this approach, we verify and synthesize 
a controller over $k$ time steps. We then compute a completeness threshold, $\overline{k}$,
for this controller, and verify the controller correctness to $\overline{k}$ time steps.
Essentially, $\overline{k}$ represents the number of iterations 
required to sufficiently unwind the closed-loop state-space system 
such that the boundaries are not violated for any $k{>}\overline{k}$
(i.e. it is sufficient to unwind the closed-loop state-space system up to  
$\overline{k}$ in order to ensure that $\phi_{safety}$ holds).

@@

%\todo{please eventually itemize next}
\begin{enumerate}
\item The inductive synthesis phase (i.e. {\sc synthesize}) uses BMC to
compute a candidate solution $K$
that satisfies both the stability criteria detailed in 
Sec.~\ref{ssec:stability} and the safety specification in Sec.~\ref{ssec:safety}.
To synthesize a controller that satisfies the stability criteria, we require that
a computed characteristic polynomial satisifies Jury's criteria. The details of this
calculation can be found in the Appendix.

%% The stability criteria are satisfied if the characteristic polynomial of $A_d - B_d * K_d$ meets Jury's criteria. This is 
%% independent of the initial states. 
Regarding the latter, we synthesize a safe controller by unwinding the transition system
$k$ steps and picking a controller $K$ and an initial state such that the states at each step do
not violate the safety criteria. This is sound if the current $k$ is below the
completeness threshold, which we assume to be true in the {\sc synthesize} phase.
We also assume some
precision $\langle I_p,F_p\rangle$ for the plant and a sampling rate.  The
checks that all these assumptions are sound are performed by subsequent
{\sc verify} stages.

@@

\item The first {\sc verify} stage, {\sc safety}, checks that the candidate
solution $K$ is safe for all possible initial states, i.e., does not reach an unsafe
state within $k$ steps where again we assume $k$ to be under the completeness threshold. 
%The safety criteria here must be valid for all possible allowed initial states, and so we
After unwinding the transition system corresponding to the previously synthesized controller
$k$ steps, we check that the safety specification holds for any initial state.

@@

In the subsequent iteration, the synthesizer is unable to find further 
suitbale candidates and it returns UNSAT, meaning that the current precision is
insufficient.  Consequently, we increase the precision to $I_p=17$,
$F_p=7$.
%
Since the previous counterexamples were obtained at lower precision,
we remove them from the set of counterexamples.  Back in the {\sc
  synthesize} phase, we re-start the process with a candidate solution
with all coefficients $0$.  Next, the {\sc safety} verification stage
provides the first counterexample at higher precision, $x_0=[-0.5
  \ 0.5 \ 0.5]$ and {\sc synthesize} finds $K=[0 \ 0.01171875
  \ 0.015625]^T$ as a candidate that eliminates this counterexample.
However, this candidate triggers the counterexample
$x_0=[0.5\ -0.5\ -0.5]$ found again by the {\sc safety} verification
stage.  In the next iteration, we get the candidate $K=[0 \ 0
  \ -0.015625]$, followed by the counterexample $x_0 = [0.5 \ 0.5
  \ 0.5]$. Finally, {\sc synthesize} finds the candidate $K=[0.01171875
  \ -0.013671875 \ -0.013671875]$, which is validated as a final
solution by all verification stages.

@@

\item In this phase we synthesize a candidate controller

@@

%-------------------------------
\subsection{Threats to validity}
%-------------------------------
We report an assessment of both our approaches over a diverse set of real-world benchmarks. Nevertheless, this set of benchmarks is limited within the scope of this paper and the performance may not generalize to other benchmarks.

@@

The definition of asymptotic stability is that the system converges to a reference signal, in this case we use no
reference signal so an asymptotically stable system will converge to the origin. We know that the original system
with an infinite precision controller is stable, because we have synthesized it to meet Jury's criteria. Hence
$(A_d  - B_dK)^n*x_0$ must converge to zero.
